---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# restrictedROC

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/restrictedROC)](https://CRAN.R-project.org/package=restrictedROC)
[![R-CMD-check](https://github.com/ggrlab/restrictedROC/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ggrlab/restrictedROC/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/ggrlab/restrictedROC/branch/main/graph/badge.svg)](https://app.codecov.io/gh/ggrlab/restrictedROC?branch=main)
<!-- badges: end -->

The goal of restrictedROC is to ...

## Installation

You can install the development version of restrictedROC like so:

``` r
devtools::install_github("ggrlab/restrictedROC")
```

## Quickstart

Given a dataset with one dependent (outcome, binary) variable and one independent 
(predictor, numeric) variables, e.g. aSAH from the pROC package:

```r
data(aSAH, package="pROC")
head(aSAH)
#    gos6 outcome gender age wfns s100b  ndka
# 29    5    Good Female  42    1  0.13  3.01
# 30    5    Good Female  37    1  0.14  8.54
# 31    5    Good Female  42    1  0.10  8.09
# 32    5    Good Female  27    1  0.04 10.42
# 33    1    Poor Female  42    3  0.13 17.40
# 34    1    Poor   Male  48    2  0.10 12.75
```
Here we use ``outcome`` and ``ndka`` as dependent and independent variables, respectively. 
``outcome`` is a factor with two levels, ``ndka`` is a numeric variable.

The most convenient usage of restrictedROC is via the ``simple_rROC_permutation()`` function.
The most important results of 
``restrictedROC::simple_rROC_permutation()`` are:

    - (restricted) AUCs for all possible cutoffs of the predictor variable
    - A global and maximal (including all restrictions) AUC
    - Permutation p-values for the global and maximal AUC


```r
# library(restrictedROC)
set.seed(412)
res_rroc <- simple_rROC_permutation(
    response = aSAH$outcome,
    predictor = aSAH$ndka,
    positive_label = "Good",
    n_permutations=100  # increase that in real data!
)
```

``res_rroc$permutation_pval`` refers to the permutation p-values of:
    - ``pval.twoside.global``: The global AUC when all samples are used. This is the usually known AUC with a calculated permutation p-value.
    - ``pval.twoside.max``: This is the maximal (restricted or unrestricted) AUC's permutation p-value. 

```r
print(res_rroc$permutation_pval)
#    pval.twoside.max pval.twoside.global      n_permutations 
#          0.16831683          0.01980198        100.00000000 
```
In this particular example we see that the global AUC is significant, but the maximal 
AUC is not. This tells that the data should **not** be restricted. 

However, if the p-value of the maximal AUC is significant, we can restrict the data
to the informative range by corresponding to the following information:


``res_rroc$global`` refers to the AUC, its variance under H0, the standardized AUC, 
and its (asymptotic, not permutation!) p-value.
```r
print(res_rroc$global)
#        auc auc_var_H0     rzAUC pval_asym
# 1 0.388042   0.388042 -1.973565 0.0484312
```
``res_rroc$max_total`` refers to the AUC, its variance under H0, the standardized AUC, 
its (asymptotic, not permutation!) p-value, the threshold(=restriction value) and which 
part of the data is **kept** and therefore **within the informative range**. 
```r
print(res_rroc$max_total)
#         auc  auc_var_H0     rzAUC   pval_asym threshold part
# 1 0.3397963 0.003492623 -2.710795 0.006712216     5.685 high
```
In this example, the maximal AUC is 0.3398, with a restriction value of 5.685 and a 
focus on the "high" part. Therefore, the informative range is ``ndka > 5.685``.


We have a convenient way to visualize the data and the results:
```r
grouped_data <- split(aSAH$ndka, aSAH$outcome)
pdf("example.pdf", width=10, height=5)
restrictedROC::plot_density_rROC_empirical(
    values_grouped=grouped_data
)
```





You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.


# How was this package created?

```r

# for VSCode
install.packages("languageserver")
install.packages("devtools")
usethis::create_tidy_package("/home/gugl/clonedgit/ggrlab/restrictedROC")
usethis::proj_activate("/home/gugl/clonedgit/ggrlab/restrictedROC")
usethis::use_tidy_style(strict = TRUE)
usethis::use_git()

```


``usethis`` tells you to envoke further github-related commands. 
There is two ways to continue: 
1. Create a personal access token (PAT) and use it to authenticate with github
2. Manually push the package to github 

Pushing manually works fine, but some advanced ``usethis`` commands won't work properly, 
therefore I will continue with the PAT.

```{r}
#
usethis::create_github_token()
gitcreds::gitcreds_set() # Then enter the freshly generated token
usethis::use_github(
    organisation = "ggrlab",
    private = FALSE,
    visibility = "public"
)
```
WARNING!!!
If an error occurs because e.g. the repository exists already at github, use the following instead of 
``usethis::use_github()``:
```bash
git branch -M main_devel
git remote add origin git@github.com:ggrlab/restrictedROC.git
git checkout main
git merge main_devel --allow-unrelated-histories
git push -u origin main
git branch --delete main_devel
```

```r
usethis::use_tidy_github()
usethis::use_tidy_github_actions()
# overwrite tidy's default "check-full" with "check-standard"
# to not run so many checks
usethis::use_github_action("check-standard")
usethis::use_tidy_github_labels()
usethis::use_pkgdown_github_pages()
```


Additional information: 
```r
usethis::use_author(
    given="Gunther", 
    family="Glehr", 
    email="gunthergl@gmx.net", 
    role=c("aut", "cre"), 
    comment = c("ORCID"="0000-0002-1495-9162")
    )
usethis::use_lifecycle()
usethis::use_news_md()
lintr::use_lintr(type = "tidyverse")
# Change manually to: 
# linters: linters_with_defaults(line_length_linter = line_length_linter(120),indentation_linter = indentation_linter(4)) # see vignette("lintr")
# encoding: "UTF-8"
```
precommit is a wonderful tool to check your code before committing it. 

```r
# https://lorenzwalthert.github.io/precommit/articles/precommit.html
# install.packages("precommit")
# bash::$ conda deactivate
# bash::$ pip3 install pre-commit
precommit::install_precommit()
precommit::use_precommit()
```
Before committing: ``pre-commit install --hook-type pre-push``, then commit. 

Used packages:
```r
usethis::use_package("ggplot2")
usethis::use_package("patchwork")
usethis::use_package("pROC")
usethis::use_package("ggpubr")
usethis::use_package("stats")
usethis::use_package("tibble")
usethis::use_package("dplyr")
usethis::use_package("labeling")
usethis::use_package("tidyr")
usethis::use_package("statmod")
usethis::use_package("future.apply")
precommit::snippet_generate('additional-deps-roxygenize')


```



# How to add information

```r

# Use pre-commits to check your code before committing it
remotes::install_github("lorenzwalthert/precommit")
precommit::install_precommit()
precommit::use_precommit()

# Increase the versions, manually or by using usethis::use_version()
usethis::use_version("minor")
usethis::use_version("dev")

# Add new functionality and document it
# During development, have a clean R environment and run devtools::load_all() to load the current status of the package
devtools::load_all()

# After adding new functionality, run devtools::check() to update the documentation
devtools::check()
devtools::document()

# Vignettes are a great way to document your package
# Add a new vignette by running 
usethis::use_vignette("vignette_name")
# Change the vignette in vignettes/vignette_name.Rmd
# Build the vignette by running
devtools::build_vignettes() # This also installs the package

devtools::build_readme() # This updates the README.md file from the README.Rmd
```

- Disable pre-commit for a single commit: ``git commit . -m 'quick fix' --no-verify``


# Troubleshooting: 

## ``devtools::build_vignettes`` fails
```
--- re-building 'first_function.Rmd' using rmarkdown
Error: processing vignette 'first_function.Rmd' failed with diagnostics:
unused argument (resolve_symlink = FALSE)
--- failed re-building 'first_function.Rmd'

SUMMARY: processing the following file failed:
  'first_function.Rmd'
```

Solution: 

```
install.packages("xfun")
```

## Vignette xfun::isFalse() will be deprecated

What:    ``devtools::check()`` throws the following error:

```
> checking re-building of vignette outputs ... ERROR
  Error(s) in re-building vignettes:
    ...
  --- re-building 'first_function.Rmd' using rmarkdown
  Quitting from lines 11-15 (first_function.Rmd) 
  Error: processing vignette 'first_function.Rmd' failed with diagnostics:
  The function xfun::isFALSE() will be deprecated in the future. Please consider using base::isFALSE(x) or identical(x, FALSE) instead.
  --- failed re-building 'first_function.Rmd'
  
  SUMMARY: processing the following file failed:
    'first_function.Rmd'
  
  Error: Vignette re-building failed.
  Execution halted
```
    
Solution: See https://stackoverflow.com/questions/76081732/problem-when-converting-r-file-to-r-markdown

```
install.packages("knitr")
```
