---
title: "AUROC vs Mann-Whitney U"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
options(warn = 1)
library(restrictedROC)
library(MASS) # for rnbinom
```
From <cite>[Wackerly et al, 2014, Cengage Learning][1]</cite>: The Mann-Whitney U test is a nonparametric test obtained by:

1. Ordering all (n_1 + n_2) observations by their value 
2. Count how many observations of sample I precede _each_ observation of sample II 
3. Sum all counts from step 2

```{r, Example Mann-Whitney U}
# Example from Wackerly 2014, p758, 15.6
values <- c(25, 26, 27, 28, 29, 31, 32, 35)
names(values) <- c("I", "I", "I", "II", "II", "I", "II", "II")
print(values)

# Count how many observations in I and II
n_1 <- length(values[names(values) == "I"])
n_2 <- length(values[names(values) == "II"])
```

```{r, U calculation}
# 1. Ordering all (n_1 + n_2) observations by their value
values_sorted <- sort(values)
print(values_sorted)

# 2. Count how many observations of sample I precede each observation of sample II
u_values <- c()
for (y_observation in values_sorted[names(values_sorted) == "II"]) {
    u_values <- c(u_values, sum(values_sorted[names(values_sorted) == "I"] < y_observation))
}

u_total <- sum(u_values)
print(u_total)
```

From <cite>[Wackerly et al, 2014, Cengage Learning][1], 15.6</cite> we see that the U statistic has a direct connection to the Wilcoxon rank sum statistic.  For that, we need to calculate the ranks of the values.
```{r, Rank sum calculation}
values_ranks <- rank(values)
print(values_ranks)

# Sum of ranks of for sample I and II
rs_I <- sum(values_ranks[names(values_ranks) == "I"])
rs_II <- sum(values_ranks[names(values_ranks) == "II"])
ranksum <- min(rs_I, rs_II)
print(ranksum)
```

```{r, Comparison to wilcoxon test}
# R's wilcox.test:
#   STATISTIC <- c(W = sum(r[seq_along(x)]) - n.x * (n.x + 1)/2)
# translated:
#  STATISTIC <- c("W"=rs_I - n_1*(n_1 + 1)/2)
w_test <- wilcox.test(
    x = values[names(values) == "I"],
    y = values[names(values) == "II"],
    paired = FALSE,
    alternative = "two.sided"
)
print(w_test)
print(c("W - manually calculated" = rs_I - n_1 * (n_1 + 1) / 2))
```

With the formula from Wackerly2014: 

$$
U = n_1 n_2 + \frac{n_1 (n_1 + 1)}{2} - ranksum_{I}
$$

we see that 
$$
ranksum_{I} = n_1 n_2 + \frac{n_1 (n_1 + 1)}{2} - U
$$

and therefore 

$$
W_{Rwilcox} = ranksum_{I} - \frac{n_1 (n_1 + 1)}{2} = n_1 n_2 - U
$$
```{r, From Mann-Whitney U statistic to Wilcoxon}
ranksum_I_from_u <- n_1 * n_2 + n_1 * (n_1 + 1) / 2 - u_total
w_r <- ranksum_I_from_u - n_1 * (n_1 + 1) / 2
w_r_alternative <- n_1 * n_2 - u_total

print(w_r_alternative)
print(w_r)
print(w_test)

u_from_wilcox <- n_1 * n_2 - w_test$statistic
print(u_from_wilcox)
print(u_total)
```

### AUROC is the same as Mann-Whitney U statistic
```{r, AUROC is the same as Mann-Whitney U statistic}
# Example from Wackerly 2014, p758, 15.6
values <- c(25, 26, 27, 28, 29, 31, 32, 35)
names(values) <- c("I", "I", "I", "II", "II", "I", "II", "II")
print(values)

# Count how many observations in I and II
n_1 <- length(values[names(values) == "I"])
n_2 <- length(values[names(values) == "II"])

auroc_from_u <- u_total / (n_1 * n_2)
print(auroc_from_u)

# Explicitely calculate AUC with pROC
print(pROC::roc(
    response = names(values),
    predictor = values,
    levels = c("I", "II"),
    direction = "<",
    auc = TRUE
))
```


### Bigger example of AUROC vs Mann-Whitney U statistic
```{r, Bigger example of AUROC vs Mann-Whitney U statistic}
set.seed(2)
values <- c(
    a = rnorm(50, mean = 8, sd = 1),
    b = rnorm(50, mean = 9, sd = 2)
)
names(values) <- gsub("[0-9]*", "", names(values))
n_1 <- length(values[names(values) == "a"])
n_2 <- length(values[names(values) == "b"])

big_roc <- pROC::roc(
    response = names(values),
    predictor = values,
    levels = c("a", "b"),
    direction = "<",
    auc = TRUE
)
big_wilcox <- wilcox.test(
    x = values[names(values) == "a"],
    y = values[names(values) == "b"],
    paired = FALSE,
    alternative = "two.sided"
)
big_ttest <- t.test(
    x = values[names(values) == "a"],
    y = values[names(values) == "b"],
    paired = FALSE,
    alternative = "two.sided"
)

u_from_wilcox <- n_1 * n_2 - big_wilcox$statistic
auc_from_u <- u_from_wilcox / (n_1 * n_2)
print(auc_from_u)
print(big_roc)

# Plot ROC curve
library(ggplot2)
df_split <- as.data.frame(split(values, names(values)))
plot_box <- ggplot(reshape2::melt(df_split), aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    geom_jitter(size = 1, width = .1) +
    scale_fill_manual(values = c("a" = "#008B4570", "b" = "#ee000070")) +
    ggpubr::theme_pubr()
plots_rroc <- restrictedROC::plot_density_ROC_empirical(
    values_grouped = df_split,
    xmin = min(values), xmax = max(values),
    levels = c("b", "a"),
    direction = "<"
)
scaling <- .35
# pdf("removeme.pdf", width = 30 * scaling, height = 10 * scaling)
set.seed(51) # for geom_jitter reproducibility
patchwork::wrap_plots(
    plot_box, plots_rroc[[1]] + scale_fill_manual(values = c("a" = "#008B4570", "b" = "#ee000070")), plots_rroc[[2]]
)
# dev.off()
```

```{r, RestrictedROC permutation test}
rroc_perm <- restrictedROC::simple_rROC_permutation(
    response = names(values),
    predictor = values,
    n_permutations = 1000,
    positive_label = "b",
    direction = "<",
    return_proc = TRUE
)
```

```{r, Compare tests}
# nolint start
print(big_roc)
# Call:
# roc.default(response = names(values), predictor = values, levels = c("a",     "b"), direction = "<", auc = TRUE)

# Data: values in 50 controls (names(values) a) < 50 cases (names(values) b).
# Area under the curve: 0.5592

print(big_ttest)
#         Welch Two Sample t-test
# data:  values[names(values) == "a"] and values[names(values) == "b"]
# t = -1.7942, df = 69.92, p-value = 0.07711
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -1.41436615  0.07477889
# sample estimates:
# mean of x mean of y
#  8.069138  8.738932

print(big_wilcox)
# data:  values[names(values) == "a"] and values[names(values) == "b"]
# W = 1102, p-value = 0.3092
# alternative hypothesis: true location shift is not equal to 0

print(rroc_perm$permutation_pval)
#    pval.twoside.max pval.twoside.global      n_permutations
#        6.993007e-03        3.226773e-01        1.000000e+03
print(rroc_perm$max_total)
#         auc auc_var_H0    rzAUC    pval_asym threshold part
# 1 0.7549342 0.00486568 3.654739 0.0002574438  7.381726 high

cat("Informative range established from restriction: (", round(rroc_perm$max_total[["threshold"]], 2), ", Inf)\n", sep = "")
# Informative range established from restriction: (7.38, Inf)
# nolint end

rroc_part_plots <- restrictedROC::plot_rROC_part(
    rROC_res = rroc_perm, threshold = rroc_perm$max_total[["threshold"]]
)
informative_range <- c(rroc_perm$max_total[["threshold"]])
# pdf("removeme_2.pdf", width = 30 * scaling, height = 10 * scaling)
set.seed(51) # for geom_jitter reproducibility
patchwork::wrap_plots(
    plot_box + geom_hline(yintercept = informative_range, col = "red", linewidth = 1),
    plots_rroc[[1]] +
        geom_vline(xintercept = informative_range, col = "red", linewidth = 1) +
        scale_fill_manual(values = c("a" = "#008B4570", "b" = "#ee000070")),
    plots_rroc[[2]]
)
# dev.off()
# pdf("removeme_3.pdf", height=4, width=4)
print(rroc_part_plots)
# dev.off()
```

[1]: Wackerly, Dennis, William Mendenhall, and Richard L. Scheaffer. Mathematical statistics with applications. Cengage Learning, 2014.
