---
title: "AUROC vs Mann-Whitney U"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
options(warn = 1)
library(restrictedROC)
library(MASS) # for rnbinom
main_plotname <- "res/paper/auroc_wilcoxon"
dir.create("res/paper", recursive = TRUE)
```
From <cite>[Wackerly et al, 2014, Cengage Learning][1]</cite>: The Mann-Whitney U test is a nonparametric test obtained by:

1. Ordering all (n_1 + n_2) observations by their value 
2. Count how many observations of sample I precede _each_ observation of sample II 
3. Sum all counts from step 2

```{r, Example Mann-Whitney U}
# Example from Wackerly 2014, p758, 15.6
values <- c(25, 26, 27, 28, 29, 31, 32, 35)
names(values) <- c("I", "I", "I", "II", "II", "I", "II", "II")
print(values)

# Count how many observations in I and II
n_1 <- length(values[names(values) == "I"])
n_2 <- length(values[names(values) == "II"])
```

```{r, U calculation}
# 1. Ordering all (n_1 + n_2) observations by their value
values_sorted <- sort(values)
print(values_sorted)

# 2. Count how many observations of sample I precede each observation of sample II
u_values <- c()
for (y_observation in values_sorted[names(values_sorted) == "II"]) {
    u_values <- c(u_values, sum(values_sorted[names(values_sorted) == "I"] < y_observation))
}

u_total <- sum(u_values)
print(u_total)
```

From <cite>[Wackerly et al, 2014, Cengage Learning][1], 15.6</cite> we see that the U statistic has a direct connection to the Wilcoxon rank sum statistic.  For that, we need to calculate the ranks of the values.
```{r, Rank sum calculation}
values_ranks <- rank(values)
print(values_ranks)

# Sum of ranks of for sample I and II
rs_I <- sum(values_ranks[names(values_ranks) == "I"])
rs_II <- sum(values_ranks[names(values_ranks) == "II"])
ranksum <- min(rs_I, rs_II)
print(ranksum)
```

```{r, Comparison to wilcoxon test}
# R's wilcox.test:
#   STATISTIC <- c(W = sum(r[seq_along(x)]) - n.x * (n.x + 1)/2)
# translated:
#  STATISTIC <- c("W"=rs_I - n_1*(n_1 + 1)/2)
w_test <- wilcox.test(
    x = values[names(values) == "I"],
    y = values[names(values) == "II"],
    paired = FALSE,
    alternative = "two.sided"
)
print(w_test)
print(c("W - manually calculated" = rs_I - n_1 * (n_1 + 1) / 2))
```

With the formula from Wackerly2014: 

$$
U = n_1 n_2 + \frac{n_1 (n_1 + 1)}{2} - ranksum_{I}
$$

we see that 
$$
ranksum_{I} = n_1 n_2 + \frac{n_1 (n_1 + 1)}{2} - U
$$

and therefore 

$$
W_{Rwilcox} = ranksum_{I} - \frac{n_1 (n_1 + 1)}{2} = n_1 n_2 - U
$$
```{r, From Mann-Whitney U statistic to Wilcoxon}
ranksum_I_from_u <- n_1 * n_2 + n_1 * (n_1 + 1) / 2 - u_total
w_r <- ranksum_I_from_u - n_1 * (n_1 + 1) / 2
w_r_alternative <- n_1 * n_2 - u_total

print(w_r_alternative)
print(w_r)
print(w_test)

u_from_wilcox <- n_1 * n_2 - w_test$statistic
print(u_from_wilcox)
print(u_total)
```

### AUROC is the equivalent to Mann-Whitney U statistic
```{r, AUROC is the equivalent to Mann-Whitney U statistic}
# Example from Wackerly 2014, p758, 15.6
values <- c(25, 26, 27, 28, 29, 31, 32, 35)
names(values) <- c("I", "I", "I", "II", "II", "I", "II", "II")
print(values)

# Count how many observations in I and II
n_1 <- length(values[names(values) == "I"])
n_2 <- length(values[names(values) == "II"])

auroc_from_u <- u_total / (n_1 * n_2)
print(auroc_from_u)

# Explicitely calculate AUC with pROC
print(pROC::roc(
    response = names(values),
    predictor = values,
    levels = c("I", "II"),
    direction = "<",
    auc = TRUE
))
```


### Bigger example of AUROC vs Mann-Whitney U statistic
```{r, Bigger example of AUROC vs Mann-Whitney U statistic}
set.seed(2)
values <- c(
    a = rnorm(50, mean = 8, sd = 1),
    b = rnorm(50, mean = 9, sd = 2)
)
names(values) <- gsub("[0-9]*", "", names(values))
n_1 <- length(values[names(values) == "a"])
n_2 <- length(values[names(values) == "b"])

big_roc <- pROC::roc(
    response = names(values),
    predictor = values,
    levels = c("a", "b"),
    direction = "<",
    auc = TRUE
)
big_wilcox <- wilcox.test(
    x = values[names(values) == "a"],
    y = values[names(values) == "b"],
    paired = FALSE,
    alternative = "two.sided"
)
big_ttest <- t.test(
    x = values[names(values) == "a"],
    y = values[names(values) == "b"],
    paired = FALSE,
    alternative = "two.sided"
)

u_from_wilcox <- n_1 * n_2 - big_wilcox$statistic
auc_from_u <- u_from_wilcox / (n_1 * n_2)
print(auc_from_u)
print(big_roc)

# Plot ROC curve
library(ggplot2)
df_split <- as.data.frame(split(values, names(values)))
plot_box <- ggplot(reshape2::melt(df_split), aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    geom_jitter(size = 1, width = .1) +
    scale_fill_manual(values = c("a" = "#ee000070", "b" = "#008B4570")) +
    ggpubr::theme_pubr()
plots_rroc <- restrictedROC::plot_density_ROC_empirical(
    values_grouped = df_split,
    xmin = min(values), xmax = max(values),
    levels = c("a", "b"),
    direction = "<"
)
scaling <- .35
pdf(paste0(main_plotname, "boxplot_density_roc.pdf"), width = 30 * scaling, height = 10 * scaling)
set.seed(51) # for geom_jitter reproducibility
patchwork::wrap_plots(
    plot_box,
    plots_rroc[[1]] +
        scale_fill_manual(values = c("a" = "#ee000070", "b" = "#008B4570")),
    plots_rroc[[2]]
)
dev.off()
```

```{r, RestrictedROC permutation test}
rroc_perm <- restrictedROC::simple_rROC_permutation(
    response = names(values),
    predictor = values,
    n_permutations = 1000,
    positive_label = "b",
    direction = "<",
    return_proc = TRUE
)
```

```{r, Compare tests}
# nolint start
print(big_roc)
# Call:
# roc.default(response = names(values), predictor = values, levels = c("a",     "b"), direction = "<", auc = TRUE)

# Data: values in 50 controls (names(values) a) < 50 cases (names(values) b).
# Area under the curve: 0.5592

print(big_ttest)
#         Welch Two Sample t-test
# data:  values[names(values) == "a"] and values[names(values) == "b"]
# t = -1.7942, df = 69.92, p-value = 0.07711
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -1.41436615  0.07477889
# sample estimates:
# mean of x mean of y
#  8.069138  8.738932

print(big_wilcox)
# data:  values[names(values) == "a"] and values[names(values) == "b"]
# W = 1102, p-value = 0.3092
# alternative hypothesis: true location shift is not equal to 0

print(rroc_perm$permutation_pval)
#    pval.twoside.max pval.twoside.global      n_permutations
#        6.993007e-03        3.226773e-01        1.000000e+03
print(rroc_perm$max_total)
#         auc auc_var_H0    rzAUC    pval_asym threshold part
# 1 0.7549342 0.00486568 3.654739 0.0002574438  7.381726 high

cat("Informative range established from restriction: (", round(rroc_perm$max_total[["threshold"]], 2), ", Inf)\n", sep = "")
# Informative range established from restriction: (7.38, Inf)
# nolint end

rroc_part_plots <- restrictedROC::plot_rROC_part(
    rroc_perm,
    threshold = rroc_perm$max_total[["threshold"]]
)
informative_range <- c(rroc_perm$max_total[["threshold"]])
pdf(paste0(main_plotname, "boxplot_density_roc_restriction.pdf"), width = 30 * scaling, height = 10 * scaling)
set.seed(51) # for geom_jitter reproducibility
patchwork::wrap_plots(
    plot_box + geom_hline(yintercept = informative_range, col = "red", linewidth = 1),
    plots_rroc[[1]] +
        geom_vline(xintercept = informative_range, col = "red", linewidth = 1) +
        scale_fill_manual(values = c("a" = "#ee000070", "b" = "#008B4570")),
    plots_rroc[[2]]
)
dev.off()
pdf(paste0(main_plotname, "boxplot_density_roc_restriction_parts.pdf"), height = 4, width = 4)
print(rroc_part_plots)
dev.off()
```

R's two-sample ``wilcoxon.test`` is (by default) calculated "exact" if there are less than 50 observations in both groups. ``n.x`` is the number of samples in the first group, ``n.y`` the number of samples in the second group. 

See https://github.com/SurajGupta/r-source/blob/master/src/library/stats/R/wilcox.test.R for the code. Briefly going through that code: 

 - Line 245 starts the two sample case 
 - Line 252: If ``exact == NULL`` (default), then `` exact <- (n.x < 50) && (n.y < 50)``
 - With ties: z-standardization approximization
 - Without ties: Exact calculation

```{r, eval = FALSE, z-standardization approximization}
r <- rank(c(x - mu, y))
STATISTIC <- c("W" = sum(r[seq_along(x)]) - n.x * (n.x + 1) / 2)

NTIES <- table(r)
z <- STATISTIC - n.x * n.y / 2
SIGMA <- sqrt((n.x * n.y / 12) * ((n.x + n.y + 1) - sum(NTIES^3 - NTIES) / ((n.x + n.y) * (n.x + n.y - 1))))
z <- (z - CORRECTION) / SIGMA

if (correct) {
    CORRECTION <- switch(alternative,
        "two.sided" = sign(z) * 0.5,
        "greater" = 0.5,
        "less" = -0.5
    )
    METHOD <- paste(METHOD, "with continuity correction")
}

PVAL <- switch(alternative,
    "less" = pnorm(z),
    "greater" = pnorm(z, lower.tail = FALSE),
    "two.sided" = 2 * min(
        pnorm(z),
        pnorm(z, lower.tail = FALSE)
    )
)
```

```{r, eval = FALSE, Exact calculation WITHOUT ties}
r <- rank(c(x - mu, y))
STATISTIC <- c("W" = sum(r[seq_along(x)]) - n.x * (n.x + 1) / 2)

PVAL <- switch(alternative,
    "two.sided" = {
        p <- if (STATISTIC > (n.x * n.y / 2)) {
            pwilcox(STATISTIC - 1, n.x, n.y, lower.tail = FALSE)
        } else {
            pwilcox(STATISTIC, n.x, n.y)
        }
        min(2 * p, 1)
    },
    "greater" = {
        pwilcox(STATISTIC - 1, n.x, n.y, lower.tail = FALSE)
    },
    "less" = pwilcox(STATISTIC, n.x, n.y)
)
```



```{r, Compare rzAUC approximation and wilcoxon statistic}
set.seed(4)
values <- c(
    a = rnorm(55, mean = 8, sd = 1),
    b = rnorm(55, mean = 8, sd = 1)
    # b = rnorm(75, mean = 8, sd = 2)
)
names(values) <- gsub("[0-9]*", "", names(values))
rroc_perm_2 <- restrictedROC::rROC(
    y = names(values),
    x = values,
    n_permutations = 0,
    positive_label = "b",
    direction = "<",
    return_proc = TRUE
)
rroc_perfs <- rroc_perm_2[[1]][[1]][["permutation"]][["performances"]]

data_df <- tibble::tibble(
    response = names(values),
    predictor = values
)
wilcox_res <- lapply(rroc_perfs[["threshold"]], function(t_x) {
    samples_above_threshold <- data_df |> dplyr::filter(predictor > t_x)
    samples_a <- samples_above_threshold |> dplyr::filter(response == "a")
    samples_b <- samples_above_threshold |> dplyr::filter(response == "b")
    if (nrow(samples_b) < 1 || nrow(samples_a) < 1) {
        return(NULL)
    }
    tmp <- wilcox.test(
        x = samples_a |> dplyr::pull(predictor),
        y = samples_b |> dplyr::pull(predictor),
        alternative = "two.sided"
    )
    return(tmp)
})
w_statistic <- unlist(lapply(wilcox_res, function(x) x$statistic))
w_pvalue <- unlist(lapply(wilcox_res, function(x) x$p.value))

rroc_perfs[["w.statistic"]] <- NA
rroc_perfs[["w.pvalue"]] <- NA
rroc_perfs[seq_along(w_statistic), "w.statistic"] <- w_statistic
rroc_perfs[seq_along(w_pvalue), "w.pvalue"] <- w_pvalue

rroc_perfs[["u_from_wilcox"]] <- with(rroc_perfs, positives_high * negatives_high - w.statistic)
rroc_perfs[["auroc_from_u"]] <- with(rroc_perfs, u_from_wilcox / (positives_high * negatives_high))

rroc_perfs[, c(1:8, 20:25)]

rroc_perfs[["fpr_global"]]
rroc_perfs[["w.pvalue"]] - rroc_perfs[["pval_asym_high"]]
pdf(paste0(main_plotname, "rzAUC_vs_wilcox.pdf"), width = 6, height = 4)
print(
    ggplot(rroc_perfs, aes(x = auc_high, y = round(auroc_from_u - auc_high, 8))) +
        geom_point() +
        ggtitle(
            "Restricted rAUC == (rAUC from wilcoxon test)?",
            subtitle = "Sanity check if restricted AUC is equivalent to calculated AUC \nfrom multiple wilcoxon test on the corresponding sample subsets"
        ) +
        ggpubr::theme_pubr() +
        theme(
            plot.title = element_text(size = 6),
            plot.subtitle = element_text(size = 6)
        )
)
print(
    ggplot(
        rroc_perfs |>
            dplyr::select(fpr_global, pval_asym_high, w.pvalue) |>
            tidyr::pivot_longer(-fpr_global, names_to = "name", values_to = "p.value"),
        aes(x = fpr_global, y = (p.value), col = name)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs |>
            dplyr::select(fpr_global, pval_asym_high, w.pvalue) |>
            tidyr::pivot_longer(-fpr_global, names_to = "name", values_to = "p.value"),
        aes(x = fpr_global, y = (p.value), col = name)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs |>
            dplyr::select(fpr_global, pval_asym_high, w.pvalue) |>
            tidyr::pivot_longer(-fpr_global, names_to = "name", values_to = "p.value"),
        aes(x = fpr_global, y = (p.value), col = name)
    ) +
        geom_point() +
        geom_point(data = rroc_perfs[, c("fpr_global", "rzAUC_high")], aes(y = -abs(rzAUC_high), col="black")) +
        ggpubr::theme_pubr()
)

print(
    ggplot(
        rroc_perfs |>
            dplyr::select(fpr_global, pval_asym_high, w.pvalue) |>
            tidyr::pivot_longer(-fpr_global, names_to = "name", values_to = "p.value"),
        aes(x = fpr_global, y = (p.value), col = name)
    ) +
        geom_line() + 
        geom_line(data = rroc_perfs[, c("fpr_global", "rzAUC_high")], aes(y = -abs(rzAUC_high), col="black")) +
        ggpubr::theme_pubr()
)

print(
    ggplot(
        rroc_perfs,
        aes(x = abs(rzAUC_high), y = pval_asym_high)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs,
        aes(x = abs(rzAUC_high), y = w.pvalue)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs,
        aes(x = fpr_global, y = w.pvalue)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs,
        aes(x = fpr_global, y = pval_asym_high)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs,
        aes(x = fpr_global, y = rzAUC_high)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)
print(
    ggplot(
        rroc_perfs,
        aes(x = pval_asym_high, y = w.pvalue)
    ) +
        geom_point() +
        ggpubr::theme_pubr()
)

dup_labels <- function(x) {
    tmp <- apply(rroc_perfs[, c("positives_high", "negatives_high")], 1, sum, na.rm = TRUE)
    tmp[sapply(x, function(y) {
        which.min(abs(rroc_perfs[["fpr_global"]] - y))
    })]
}
print(
    ggplot(
        rroc_perfs,
        aes(x = fpr_global, y = pval_asym_high - w.pvalue)
    ) +
        geom_point() +
        ggpubr::theme_pubr() +
        scale_x_continuous(sec.axis = dup_axis(labels = dup_labels, name = "Total number of samples"))
)
dev.off()
```

[1]: Wackerly, Dennis, William Mendenhall, and Richard L. Scheaffer. Mathematical statistics with applications. Cengage Learning, 2014.
