---
title: "Datatypes 01: Proteomics"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

In <cite>[Harel et al, 2019, Cell][1]</cite>, the authors describe melanoma response to immunotherapy. They use high-resolution mass spectrometry to quantify 4,300 proteins across 116 patients. Formalin-fixed paraffin-embedded (FFPE) tissues from 42 metastatic melanoma patients treated with TIL-based adoptive cell transfer and 74 treated with anti-PD1 immunotherapy were included. 
Samples were taken before treatment except for two patients taken 11 or 18 days after treatment initialization. Patients are split into 21/40 responders, 21/27 non-responders and 0/7 stable disease according to RECIST v1.0 guidelines (TIL/anti-PD1 cohorts). See ``Tumor sample collection`` in the paper's method section. 

We used the data available under [Table S2](https://www.sciencedirect.com/science/article/pii/S0092867419309006?via%3Dihub#app2) of the publication in ``1-s2.0-S0092867419309006-mmc1.xlsx``, sheet ``S1C`` and ``S1D``.
According to the section ``Proteomics statistical analysis`` in the paper's method section, the following analysis was performed . 

    For the clinical tumor data, normalized ratio tumor/SILAC data were log2-transformed and the protein groups were filtered to have at least 70% valid values, reaching a list of 4588,4620 or 4416 protein groups for the TIL, the anti-PD1 or the combined datasets respectively, which were further used for all downstream analyses. Dataset integration was based on gene name; multiple entries for the same gene name were integrated to a single entry by calculating the median expression value. Data were normalized by subtracting most frequent value in each sample. To extract DEPs between responders and non-responders multiple comparisons were performed; a two-sample Student's t test was performed with a p value threshold of 0.05 for each of the two datasets, following by 2D annotation enrichment test (Cox and Mann, 2012) with FDR q-value 0.02 to derive the differential functional groups in each of the two datasets. A two-sample Student's t test with a permutation-based FDR q-value of 0.1 and S0 (Tusher et al., 2001) of 0.1 was performed for the anti-PD1 dataset or the combined dataset of both treatments, to extract significantly changing proteins. 

From main text; ``Functional Analysis of Responders and Non-responders to Immunotherapy``:

    Initial bioinformatic analysis examined the functional differences between responders and non-responders in each therapy. We first applied a low stringency Student's t test with a nominal p value cutoff (p value < 0.05) and found 414 and 636 differentially expressed proteins (DEPs) in the TIL and the anti-PD1 cohorts [...].



```{r setup}
options(warn = 1)
library(restrictedROC)
```

```{r download data}
library(dplyr)
dir.create("intermediate_data")
download.file(
    url = "https://ars.els-cdn.com/content/image/1-s2.0-S0092867419309006-mmc1.xlsx",
    destfile = "intermediate_data/harel2019.xlsx",
)
harel_pheno <- readxl::read_excel("intermediate_data/harel2019.xlsx", sheet = "S1A", skip = 1, na = "NaN")
# Patients were categorized into responders (complete or partial regression) and
# non-responders (progressive disease) according to RECIST v1.0 guidelines
# "Response" contains "PD", "PR", "CR" or "SD" (stable disease)
harel_pheno <- harel_pheno |>
    # Filter out stable disease
    dplyr::filter(Response %in% c("PD", "PR", "CR")) |>
    # Add a column with the response as defined in the paper
    dplyr::mutate(response_paper = ifelse(Response == "PD", "non-responder", "responder"))
outcome <- harel_pheno[["response_paper"]]
names(outcome) <- harel_pheno[["Sample ID"]]

# S1C and S1D are the features with at least 70% present values
harel_til <- readxl::read_excel("intermediate_data/harel2019.xlsx", sheet = "S1C", skip = 1, na = "NaN")
harel_aPD1 <- readxl::read_excel("intermediate_data/harel2019.xlsx", sheet = "S1D", skip = 1, na = "NaN")

# The description columns start with a T
harel_til_mat <- as.matrix(harel_til[, !grepl("^T:", colnames(harel_til))])
# rownames(harel_til_mat) <- harel_til[["T: Gene names"]]
rownames(harel_til_mat) <- harel_til[["T: Protein IDs"]]
if (anyDuplicated(rownames(harel_til_mat)) != 0) {
    stop()
}
# rownames(harel_til_mat) <- sub(";.*", "", rownames(harel_til_mat))

harel_aPD1_mat <- as.matrix(harel_aPD1[, !grepl("^T:", colnames(harel_aPD1))])
rownames(harel_aPD1_mat) <- harel_aPD1[["T: T: Gene name"]]
if (anyDuplicated(rownames(harel_aPD1_mat)) != 0) {
    stop()
}
# rownames(harel_aPD1_mat) <- sub(";.*", "", rownames(harel_aPD1_mat))
```

```{r Reproduce t-tests}
harel_ttests <- lapply(list("til" = harel_til_mat, "aPD1" = harel_aPD1_mat), function(datapart_x) {
    res_t <- apply(datapart_x, 1, function(feature_x) {
        # names(x) are the column names of the matrix
        # The response needs to be ordered according to the column names
        x_NonResponder <- split(feature_x, outcome[names(feature_x)])
        res_test <- t.test(
            x_NonResponder[["responder"]],
            x_NonResponder[["non-responder"]],
            alternative = "two.sided", paired = FALSE, var.equal = FALSE
        )
        return(c(
            "neglog_p.value" = -log10(res_test$p.value),
            "p.value" = res_test$p.value,
            "t" = res_test$statistic[[1]],
            "diff_nr_r" = -(res_test$estimate["mean of y"] - res_test$estimate["mean of x"])[[1]]
        ))
    })
    return(tibble::as_tibble(t(res_t), rownames = "gene") |> dplyr::arrange(diff_nr_r))
})
harel_ttests
lapply(harel_ttests, function(x) sum(x$p.value < 0.05, na.rm = TRUE))
#
harel_til_mat_ttest <- apply(harel_til_mat, 1, function(x) {
    # names(x) are the column names of the matrix
    # The response needs to be ordered according to the column names
    x_NonResponder <- split(x, outcome[names(x)])
    res_test <- t.test(
        x_NonResponder[["responder"]],
        x_NonResponder[["non-responder"]],
        alternative = "two.sided", paired = FALSE, var.equal = FALSE
    )
    return(c("p.value" = res_test$p.value, "t" = res_test$statistic[[1]]))
})
tibble::as_tibble(t(harel_til_mat_ttest), rownames = "gene") |>
    dplyr::arrange(t)

x_NonResponder <- split(harel_til_mat["AMBP", ], outcome[colnames(harel_til_mat)])
tmp <- t.test(
    (na.omit(x_NonResponder[["responder"]])),
    (na.omit(x_NonResponder[["non-responder"]])),
    alternative = "two.sided", paired = FALSE, var.equal = TRUE
)

names(tmp)
diff(tmp$estimate)
print(tmp)
print(sum(harel_til_mat_ttest < 0.05, na.rm = TRUE))
tibble::tibble("gene" = names(harel_til_mat_ttest), "pvalue" = harel_til_mat_ttest)
```




```{r Why can I not reproduce the t-tests?}
download.file(
    url = "https://ars.els-cdn.com/content/image/1-s2.0-S0092867419309006-mmc2.xlsx",
    destfile = "intermediate_data/harel2019-S2.xlsx",
)
harel_pheno |>
    dplyr::filter(Cohort == "TIL") |>
    dplyr::pull(Response) |>
    table()
harel_til_s2_ttest <- readxl::read_excel("intermediate_data/harel2019-S2.xlsx", sheet = "S2A", skip = 1, na = "NaN")
ncol(harel_til_s2_ttest)
# https://github.com/SurajGupta/r-source/blob/master/src/library/stats/R/t.test.R

nx <- 21
ny <- 21
pval_equalvar <- function(tstat) {
    return(2 * pt(-abs(tstat), nx + ny - 2))
}

pdf("removeme.pdf")
hist(pval_equalvar(harel_til_s2_ttest[["Student's T-test Test statistic R_NR"]]))
dev.off()
t
```

[1]: Harel et al. "Proteomics of melanoma response to immunotherapy reveals mitochondrial dependence." Cell 179.1 (2019): 236-250. https://doi.org/10.1016/j.cell.2019.08.012
